A Safe and High-Performance Aliasing-Tolerant Parallel Programming Model

### 2012-11-14 doc/types-pldi2013/alt_example.tex
hopefully last wave of edits, spell check

\begin{lstlisting}[label={lst:tuples},caption={Tuples and Pointers Example}]
--                        voltage,current,charge,capacitance,piece ID
type CircuitNode        = <float,float,float,float,int>
--                      owned node, owned or ghost node, resistance, current
type CircuitWire<rn,rg>  = <CircuitNode@rn, CircuitNode@(rn,rg),float,float>
\end{lstlisting}

The most important decision in any Legion program is how to partition
program data.  An ideal partitioning depends on many factors,
including the shape of data structures, the input, and the number of
partitions, which usually varies with the target machine.  To allow
the necessary flexibility, in Legion partitioning is simply programmed,
rather than expressed in a more structured and restricted
sublanguage.  In the example we invoke METIS\cite{Metis98}, a 
third-party dynamic graph partitioning library (see Listing~\ref{lst:metis}).

@@

\begin{figure}[t]
\centering
\subfigure[Node region tree.]{
\label{sfig:part_fig:tree}
\includegraphics[scale=0.50]{figs/OwnedTree.pdf}
}
\subfigure[Circuit piece coloring.]{
\label{sfig:part_fig:pieces}
\includegraphics[scale=0.67]{figs/Owned_Nodes.pdf}
}
\subfigure[Ghost {\tt rg0} coloring.]{
\label{sfig:part_fig:ghost_zero}
\includegraphics[scale=0.67]{figs/Ghost_RN0.pdf}
}
\subfigure[Ghost {\tt rg1} coloring.]{
\label{sfig:part_fig:ghost_one}
\includegraphics[scale=0.67]{figs/Ghost_RN1.pdf}
}
\vspace{-2mm} 
\caption{Partitions of $r{\_}all{\_}nodes$. \label{fig:part_fig}}
\vspace{-6mm}
%  \caption[My caption
\end{figure}

@@

Colorings are the input to partition operations in Legion.  The partition
statement shown in Listing~\ref{lst:partition} performs the ``owned nodes''
partitioning.    Since colorings assign each region element at most one color
(dynamically checked), the partition operation results in 
{\em disjoint} subregions, each a subset of the {\em parent} region.
A partition operation introduces constraints into the static type environment describing
both the disjointness of subregions (e.g., {\tt rn0 $*$ rn1})
and the subregion relationships
(e.g., {\tt rn0 $\leq$ all\_nodes}).  A partition created from a multicoloring
only introduces the subregion relationships, as the partition may be aliased.

@@

\begin{lstlisting}[label={lst:partition},caption={Partition Operation Example}]
let piece_coloring : coloring(all_nodes) = create_piece_coloring[rl,all_nodes](node_list) in
partition all_nodes using piece_coloring as rn0,rn1... in ...
\end{lstlisting}

@@

Listing~\ref{lst:loop} shows the main time step loop run by the 
recursive function {\em execute\_time\_steps} (line 11)
as well the leaf task functions (lines 2-7).  
For each time step, the loop unpacks 
two previously packed circuit pieces, giving new names to the subregions introduced
by each region relationship.  The {\em execute\_time\_steps} function
will have read/write privileges for the newly named regions, such as {\tt rn0},
because it has read/write privileges for {\tt rn} and the {\tt CircuitPiece} 
region relationship ensures that {\tt rn0 $\leq$ rn}.

The {\tt execute\_time\_steps} function   
illustrates the importance of having different partitions provide 
multiple views onto the same logical region.  Both {\tt calc\_new\_currents} 
and {\tt distribute\_charge} (lines 2-5)
use the owned and ghost regions of a piece, which are from different partitions. In
{\tt calc\_new\_currents} these regions only need read
privileges, allowing both instances of the task to be
run in parallel.  In {\tt distribute\_charge} the
privilege is for a reduction which can also be done in parallel
because of the atomic and commutative nature of reductions.  Finally,
the {\tt update\_voltage} function (line 6) modifies only the disjoint owned regions, 
permitting each piece to be updated in parallel.  No single partition of the nodes
describes these data sharing patterns.

\lstset{
  captionpos=b,
  language=Haskell,
  basicstyle=\scriptsize,
  numbers=left,
  numberstyle=\tiny,
  columns=fullflexible,
  stepnumber=1,
  escapechar=\#,
  keepspaces=true,
  belowskip=-10pt,
  literate={<}{{$\langle$}}1 {>}{{$\rangle$}}1,
  morekeywords={function,rr,int,float,bool,isnull,partition,as,downregion,upregion,reads,writes,rdwrs,reduces,read,write,reduce,using,unpack,pack,coloring,multicoloring,color,newcolor,atomic,simultaneous},
  deletekeywords={head,min,max}
}
\begin{lstlisting}[float={t},label={lst:loop},caption={Main Simulation Loop}]
-- Leaf Task Declarations (implementations in appendix)
function calc_new_currents[rl,rw,rn,rg] ( ptr_list : WireList<rl,rw,rn,rg>@rl ), 
      reads(rl,rw,rn,rg), writes(rw) : bool
function distribute_charge[rl,rw,rn,rg] ( ptr_list : WireList<rl,rw,rn,rg>@rl ), 
      reads(rl,rw,rn), reduces(reduce_charge,rn,rg), atomic(rn,rg) : bool 
function update_voltage[rl,rn] ( ptr_list : NodeList<rl,rn>@rl ), 
      reads(rl,rn), writes(rn) : bool
-- Reduction function for distribute charge
function reduce_charge ( node : CircuitNode, current : float ) : CircuitNode

@@

The {\tt execute\_time\_steps} function also demonstrates the need to 
dynamically discover parallelism through region disjointess.  
To run the two instances of 
{\tt calc\_new\_currents} in parallel requires knowing that {\tt rw0}
and {\tt rw1} are disjoint (lines 17-18).  There is a similar requirement for parallel
execution of the two {\tt update\_voltage} calls with {\tt rn0} and {\tt rn1} (lines 21-22).  In
both cases, this knowledge was statically available when the subregions were created 
and could have been captured in a region relationship at the cost
of much more complicated code.  Instead, we use a simpler region relationship
and rely on an inexpensive dynamic disjointness
check by the Legion runtime to (re)discover the parallelism.

>>>>

\begin{lstlisting}[label={lst:tuples},caption={Tuples and Pointers Example}]
--                        voltage, current, charge, capacitance, piece ID
type CircuitNode        = <float, float, float, float, int>
--                      owned node, owned or ghost node, resistance, current
type CircuitWire<rn,rg>  = <CircuitNode@rn, CircuitNode@(rn,rg), float, float>
\end{lstlisting}

The most important decision in any Legion program is how to partition
program data.  An ideal partitioning depends on many factors,
including the shape of data structures, the input, and the desired number of
partitions (which usually varies with the target machine).  To allow
the necessary flexibility, in Legion partitioning is simply programmed,
rather than expressed in a more structured and restricted
sublanguage.  In the example we invoke METIS\cite{Metis98}, a 
third-party dynamic graph partitioning library (see Listing~\ref{lst:metis}).

@@

\begin{figure}[t]
\centering
\subfigure[Node region tree.]{
\label{sfig:part_fig:tree}
\includegraphics[scale=0.50]{figs/OwnedTree.pdf}
}
\subfigure[Circuit piece coloring.]{
\label{sfig:part_fig:pieces}
\includegraphics[scale=0.67]{figs/Owned_Nodes.pdf}
}
\subfigure[Ghost {\tt rg0} coloring.]{
\label{sfig:part_fig:ghost_zero}
\includegraphics[scale=0.67]{figs/Ghost_RN0.pdf}
}
\subfigure[Ghost {\tt rg1} coloring.]{
\label{sfig:part_fig:ghost_one}
\includegraphics[scale=0.67]{figs/Ghost_RN1.pdf}
}
\vspace{-2mm} 
\caption{Partitions of $r{\_}all{\_}nodes$. \label{fig:part_fig}}
\vspace{-4mm}
%  \caption[My caption
\end{figure}

@@

Colorings are the input to partition operations in Legion.  The partition
statement shown in Listing~\ref{lst:partition} performs the ``owned nodes''
partitioning.    Since colorings assign each region element at most one color,
the partition operation results in 
{\em disjoint} subregions, each a subset of the {\em parent} region.
A partition operation introduces constraints into the static type environment describing
both the disjointness of subregions (e.g., {\tt rn0 $*$ rn1})
and the subregion relationships
(e.g., {\tt rn0 $\leq$ all\_nodes}).  A partition created from a multicoloring
only introduces the subregion relationships, as the partition may be aliased.

@@

\begin{lstlisting}[label={lst:partition},caption={Partition Operation Example}]
let owned_map : coloring(all_nodes) = create_piece_coloring[rl,all_nodes](node_list) in
partition all_nodes using owned_map as rn0,rn1... in ...
\end{lstlisting}

@@

Listing~\ref{lst:loop} shows the main time step loop run by the 
recursive function {\em execute\_time\_steps} (lines 13-24)
as well the leaf task functions (lines 2-7).  
For each time step, the loop unpacks 
two previously packed circuit pieces, giving new names to the subregions introduced
by each region relationship.  The {\em execute\_time\_steps} function
will have read/write privileges for the newly named regions, such as {\tt rn0},
because it has read/write privileges for {\tt rn} and the {\tt CircuitPiece} 
region relationship ensures that {\tt rn0 $\leq$ rn}.

The {\tt execute\_time\_steps} function   
illustrates the importance of having different partitions provide 
multiple views onto the same logical region.  Both {\tt calc\_new\_currents} 
and {\tt distribute\_charge} (lines 2-5)
use the owned and ghost regions of a piece, which are from different partitions. In
{\tt calc\_new\_currents} these regions only need read
privileges, allowing both instances of the task to be
run in parallel.  In {\tt distribute\_charge} the
privilege is for a reduction which can also be done in parallel
because of the atomic and commutative nature of reductions.  Finally,
the {\tt update\_voltage} function (lines 6-7) modifies only the disjoint owned regions, 
permitting each piece to be updated in parallel.  No single partition of the nodes
describes these data sharing patterns.

\lstset{
  captionpos=b,
  language=Haskell,
  basicstyle=\scriptsize,
  numbers=left,
  numberstyle=\tiny,
  columns=fullflexible,
  stepnumber=1,
  escapechar=\#,
  keepspaces=true,
  belowskip=-10pt,
  literate={<}{{$\langle$}}1 {>}{{$\rangle$}}1,
  morekeywords={function,rr,int,float,bool,isnull,partition,as,downregion,upregion,reads,writes,rdwrs,reduces,read,write,reduce,using,unpack,pack,coloring,multicoloring,color,newcolor,atomic,simultaneous},
  deletekeywords={head,min,max}
}
\begin{lstlisting}[float={t},label={lst:loop},caption={Main Simulation Loop}]
-- Leaf Task Declarations (implementations in appendix)
function calc_new_currents[rl,rw,rn,rg] ( ptr_list : WireList<rl,rw,rn,rg>@rl ), 
      reads(rl,rw,rn,rg), writes(rw) : bool
function distribute_charge[rl,rw,rn,rg] ( ptr_list : WireList<rl,rw,rn,rg>@rl ), 
      reads(rl,rw,rn), reduces(reduce_charge,rn,rg), atomic(rn,rg) : bool 
function update_voltage[rl,rn] ( ptr_list : NodeList<rl,rn>@rl ), 
      reads(rl,rn), writes(rn) : bool

-- Reduction function for distribute charge
function reduce_charge ( node : CircuitNode, current : float ) : CircuitNode

@@

The {\tt execute\_time\_steps} function also demonstrates the need to 
dynamically discover parallelism through region disjointness.  
To run the two instances of 
{\tt calc\_new\_currents} in parallel requires knowing that {\tt rw0}
and {\tt rw1} are disjoint (lines 17-18).  There is a similar requirement for parallel
execution of the two {\tt update\_voltage} calls with {\tt rn0} and {\tt rn1} (lines 21-22).  In
both cases, this knowledge was statically available when the subregions were created 
and could have been captured in a region relationship at the cost
of much more complicated code.  Instead, we are able to use a simpler region relationship
and rely on an inexpensive dynamic disjointness
check by the Legion runtime to (re)discover the parallelism.

### 2012-11-14 doc/types-pldi2013/evaluation.tex
hopefully last wave of edits, spell check

%\subsubsection{Fluid Simulation}
%\label{subsec:fluid}
Fluid is a distributed memory version of the {\tt fluidanimate} benchmark from the PARSEC 
suite\cite{bienia11benchmarking}.  Fluid simulates the flow of an incompressible fluid
using particles that move within a regular grid of cells.  To perform operations in 
parallel the array of cells is partitioned.  Unlike
Circuit, Fluid creates and partitions regions before
allocating cells in them. 
%(i.e., partition-then-allocate).  
Another difference is that Fluid
maintains separate regions for ghost cells rather using multiple partitions of
the regions containing shared data.  Region relationships are used to capture
which regions are required for each grid.

@@

Dynamically creating and partitioning regions at runtime is crucial to Legion's
ability to handle applications that make runtime decisions about data organization (AMR).  
% FIXME
% I've deemphasized the point about allocate-then-partition vs. partition-then-allocate.
% If we want to put it back in it needs to be go in multiple places, not just here.
%Legion is able to capture both allocate-then-partition (Circuit) and partition-then-allocate (Fluid) algorithms. 
Having multiple partitions of regions is necessary for describing the many ways 
that data can be accessed (Circuit, AMR).  All the types of privileges and coherence 
are needed in some application;  region relationships
are used in all applications.   Finally, all applications introduce aliasing of
data through both multi-colorings or having multiple partitions.  Our implementations of
these applications both type check and execute, proving that our type system is
sufficiently expressive to handle real-world applications.
% written in the Legion programming model.
%, region relationships convey disjointness information.
% despite
%regions being a first-class feature that cannot be fully tracked statically.

@@

Figures~\ref{fig:ckt_overhead}, \ref{fig:fluid_overhead}, and \ref{fig:amr_overhead} show that
the overhead of the Legion runtime is always less than 7\% of the total execution time
of the applications.  In some applications communication does not scale well, but this is
a result of the algorithm required by the application, not the Legion runtime.
Even in the case of the Circuit application, which exhibits quadratic communication scaling, the Legion
runtime is able to achieve a 62.5X speedup on 96 GPUs over a hand-coded single GPU implementation\cite{Legion12}.

>>>>

%\subsubsection{Fluid Simulation}
%\label{subsec:fluid}
Fluid is a distributed memory version of the {\tt fluidanimate} benchmark from the PARSEC 
suite\cite{bienia11benchmarking}.  Fluid simulates the flow of an incompressible fluid
using particles that move within a regular grid of cells.  To perform operations in 
parallel, the array of cells is partitioned.  Unlike
Circuit, Fluid creates and partitions regions before
allocating cells in them. 
%(i.e., partition-then-allocate).  
Another difference is that Fluid
maintains separate regions for ghost cells rather using multiple partitions of
the regions containing shared data.  Region relationships are used to capture
which regions are required for each grid.

@@

Dynamically creating and partitioning regions at runtime is crucial to Legion's
ability to handle applications that make runtime decisions about data organization (AMR).  
% FIXME
% I've deemphasized the point about allocate-then-partition vs. partition-then-allocate.
% If we want to put it back in it needs to be go in multiple places, not just here.
%Legion is able to capture both allocate-then-partition (Circuit) and partition-then-allocate (Fluid) algorithms. 
Having multiple partitions of regions is necessary for describing the many ways 
that data can be accessed (Circuit, AMR).  All the types of privileges and coherence 
are needed in some application;  region relationships
are used in all applications.   Finally, all applications introduce aliasing of
data through both multicolorings or having multiple partitions.  Our implementations of
these applications both type check and execute, proving that our type system is
sufficiently expressive to handle real-world applications.
% written in the Legion programming model.
%, region relationships convey disjointness information.
% despite
%regions being a first-class feature that cannot be fully tracked statically.

@@

Figures~\ref{fig:ckt_overhead}, \ref{fig:fluid_overhead}, and \ref{fig:amr_overhead} show that
the overhead of the Legion runtime is always less than 7\% of the total execution time
of the applications.  In some applications communication does not scale well, but this is
a result of the algorithm required by the application, not the Legion runtime.
Even in the case of the Circuit application, which exhibits quadratic increases in communication cost, the Legion
runtime is able to achieve a 62.5X speedup on 96 GPUs over a hand-coded single GPU implementation\cite{Legion12}.

### 2012-11-14 doc/types-pldi2013/intro.tex
hopefully last wave of edits, spell check

%To guarantee the safety and correctness of programs, Legion relies on
%a combination of static and dynamic analysis based on the privilege
%and coherence systems.  
%To efficiently guarantee the safety and
%correctness of programs using first-class regions,
%privilege and coherence properties in the presence of arbitrary region
%aliasing, Legion relies on a combination of static and
%dynamic checks.  
The critical feature in Legion that makes static analysis tractable
and dynamic analysis efficient in the presence of potential aliasing is logical regions.
Alias analysis is a very hard static analysis problem, but is both easy and
inexpensive if done dynamically at the granularity of logical regions instead
of individual memory locations; similarly
checks on individual memory accesses are expensive if done dynamically, but
cheap if done statically within the context of local logical region privileges.
%A key theme that runs through the design is that 
%alias analysis is a very hard static analysis problem, but is both easy and
%inexpensive if done dynamically at the granularity of regions instead of individual
%heap locations, while checks that must be done on individual values are only
%cheap if done statically.  
Explicating and evaluating the design of Legion's
static and dynamic semantics is the topic of this paper, 
%and its implications are the topics of this paper,
which we present as a series of contributions:
\begin{itemize}
\item We present a type system for {\em Core Legion} programs that statically
  verifies the safety of individual pointers and region
  privileges at task call boundaries (Section~\ref{subsec:coretypes}).
%  A subtle issue arises in checking privileges when regions are
%  stored in and later retrieved from heap data structures.  The type
%  system provides enough static information that the Legion runtime
%  can leverage its dynamic information about region aliasing to
%  perform dynamic privilege checks guaranteeing safety.

>>>>

%To guarantee the safety and correctness of programs, Legion relies on
%a combination of static and dynamic analysis based on the privilege
%and coherence systems.  
%To efficiently guarantee the safety and
%correctness of programs using first-class regions,
%privilege and coherence properties in the presence of arbitrary region
%aliasing, Legion relies on a combination of static and
%dynamic checks.  
The critical feature in Legion that makes both static analysis tractable
and dynamic analysis efficient in the presence of potential aliasing is logical regions.
Alias analysis is a very hard static analysis problem, but is both easy and
inexpensive if done dynamically at the granularity of logical regions instead
of individual memory locations; similarly
checks on individual memory accesses are expensive if done dynamically, but
cheap if done statically within the context of local logical region privileges.
%A key theme that runs through the design is that 
%alias analysis is a very hard static analysis problem, but is both easy and
%inexpensive if done dynamically at the granularity of regions instead of individual
%heap locations, while checks that must be done on individual values are only
%cheap if done statically.  
Explicating and evaluating the design of Legion's
static and dynamic semantics is the topic of this paper, 
%and its implications are the topics of this paper,
which we present as a series of contributions:
\begin{itemize}
\item We present a type system for {\em Core Legion} programs that statically
  verifies the safety of individual pointers and region
  privileges at task call boundaries (Section~\ref{subsec:coretypes}).
%  A subtle issue arises in checking privileges when regions are
%  stored in and later retrieved from heap data structures.  The type
%  system provides enough static information that the Legion runtime
%  can leverage its dynamic information about region aliasing to
%  perform dynamic privilege checks guaranteeing safety.
